---
title: "QB Salary Prediction"
output: html_document
date: "2025-03-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Packages and Data Import
```{r}
# Packages, install if necessary
library(dplyr)
library(randomForest)
library(ggplot2)
library(ggrepel)
library(gridExtra)

# Straight from github repo, assumes all files in same directory
Salary_Data <- read.csv("./NFL Salary Data.csv")
Passing_Data <- read.csv("./Passing Data.csv")
Rushing_Data <- read.csv("./Rushing Data.csv")
```

Formatting Data
```{r}
# Format Player_Data
# Remove space in team name and format dollar amounts into integers
Salary_Data <- Salary_Data %>% 
  mutate(Team.Currently.With = gsub("\\n.+", "", Team.Currently.With)) %>%
  mutate(
    Value = as.numeric(gsub("[$,]", "", Value)),
    Average.Salary = as.numeric(gsub("[$,]", "", Average.Salary)),
    Practical.Guarantee = as.numeric(gsub("[$,]", "", Practical.Guarantee)),
    Guarantee.at.Sign = as.numeric(gsub("[$,]", "", Guarantee.at.Sign))
  )

QB_Salary <- Salary_Data %>%
  filter(Pos == "QB") %>% 
  select(Player, Average.Salary, Age.At.Signing, Start.Year) # Change Average.Salary if we want to use total contract or years

QB_Rushing_Data <- Rushing_Data %>% 
  filter(Pos == "QB") %>%
  select(Player, Year, Att, Yds, TD, Y.G)

QB_Data <- Passing_Data %>%
  left_join(QB_Rushing_Data, by= c("Player","Year"), suffix = c("_pass", "_rush")) %>% # Add in rushing data
  left_join(QB_Salary, by = "Player") %>% # Add in Salary Data
  filter(!is.na(Average.Salary)) %>% filter(!is.na(QBR)) %>% # Remove NAs
  filter(Att_pass > 200) # We need to include this to ensure no 3-throw champions are in the mix
  
nrow(QB_Data)
```

# Exploratory Data Analysis

Outlier Bounds via Quantiles
```{r warning=F} 
# Outlier testing
pd2024 <- QB_Data

quantilePlot <- function(df, col, title = col) {
  Q1 <- quantile(df[[col]], .25)
  Median <- quantile(df[[col]], .5)
  Q3 <- quantile(df[[col]], .75)
  IQR <- Q3 - Q1
  lower <- Q1 - 1.5 * IQR
  upper <- Q3 + 1.5 * IQR
  
  return(
    ggplot(df, aes(x = 0, y=.data[[col]])) +
    geom_boxplot() +
    geom_label(data = df, aes(y = Q1, label = paste("Q1:", round(Q1, 2)))) +
    geom_label(data = df, aes(y = Median, label = paste("Median:", round(Median, 2)))) +
    geom_label(data = df, aes(y = Q3, label = paste("Q3:", round(Q3, 2)))) +
    geom_text_repel(aes(label = 
      ifelse(
        .data[[col]] < lower | .data[[col]] > upper,
        paste(gsub(".+ ", "", Player), Year), ""
      )
    )) +
    labs(x = "", y = "", title = title)
  )
}

grid.arrange(
  quantilePlot(pd2024, "QBR"),
  quantilePlot(pd2024, "Rate", "Passer Rating"),
  quantilePlot(pd2024, "Y.G_pass", "Passing Yards per Game"),
  quantilePlot(pd2024, "TD_pass", "Touchdown Passes"),
  nrow=2, ncol=2, top = "Performance metric Quantiles:"
)
```

Outlier Visualization based on Bounds
```{r, warning=F}
# Contract v Age test
ggplot(QB_Salary, aes(Age.At.Signing, Average.Salary)) + geom_point() +
  # geom_text_repel(aes(label = Player)) +  # idt we neeeed names here
  theme(legend.position = "none") +
  labs(
    title = "How Age affects Contract Value",
    y="Yearly Salary signed for",
    x = "Age at Contract Signing"
  )

# QBR Outlier Test
ggplot(pd2024, aes(QBR, Average.Salary, color=Player)) + geom_point() +
  geom_text_repel(aes(label = 
    ifelse(
      (QBR > quantile(QBR, .75) & Average.Salary < 2e+07) | (QBR < quantile(QBR, .25) & Average.Salary > 2e+07),  
      paste(gsub("[^A-Z]+", ".", Player), sub("20", "", paste(Year))),
      ""
    ) 
  )) +
  labs(
    title = "QBR vs Average Salary, Overperformers and Underperformers labeled",
    y = "Average Salary"
  ) + 
  theme(legend.position = "none")

# Passer Rating outlier test
ggplot(pd2024, aes(Rate, Average.Salary, color=Player)) + geom_point() +
  geom_text_repel(aes(label = 
    ifelse(
      (Rate > quantile(Rate, .75) & Average.Salary < 2e+07) | (Rate < quantile(Rate, .25) & Average.Salary > 2e+07),  
      paste(gsub("[^A-Z]+", ".", Player), sub("20", "", paste(Year))),
      ""
    ) 
  )) +
  labs(
    title = "Passer Rating vs Average Salary, Overperformers and Underperformers labeled",
    y = "Average Salary",
    x = "Passer Rating"
  ) + 
  theme(legend.position = "none")

# Y.G Outlier test
ggplot(pd2024, aes(Y.G_pass, Average.Salary, color=Player)) + geom_point() +
  geom_text_repel(aes(label = 
    ifelse(
      (Y.G_pass > quantile(Y.G_pass, .75) & Average.Salary < 2e+07) | (Y.G_pass < quantile(Y.G_pass, .25) & Average.Salary > 2e+07),  
      paste(gsub("[^A-Z]+", ".", Player), sub("20", "", paste(Year))),
      ""
    ) 
  )) +
  labs(
    title = "Yards/Game v Average Salary, Overperformers and Underperformers labeled",
    y = "Average Salary",
    x = "Yards/Game"
  ) + 
  theme(legend.position = "none")

# Total Touchdown test
ggplot(pd2024, aes(TD_pass, Average.Salary, color=Player)) + geom_point() +
  geom_text_repel(aes(label = 
    ifelse(
      (TD_pass > quantile(TD_pass, .75) & Average.Salary < 2e+07) | (TD_pass < quantile(TD_pass, .25) & Average.Salary > 2e+07),  
      paste(gsub("[^A-Z]+", ".", Player), sub("20", "", paste(Year))),
      ""
    ) 
  )) +
  labs(
    title = "Passing TD v Average Salary, Overperformers and Underperformers labeled",
    y = "Average Salary",
    x = "Passing Touchdowns"
  ) + 
  theme(legend.position = "none")
```

# Training the Model

Training Data Generation
```{r}
# Use these to get static values to remove outliers with
QBR_q1 <- quantile(QB_Data$QBR , .25)
QBR_q3 <- quantile(QB_Data$QBR , .75)
Rtg_q1 <- quantile(QB_Data$Rate, .25)
Rtg_q3 <- quantile(QB_Data$Rate, .75)


possible_predictors <- QB_Data[,c(3, 6, 7, 9:18, 20:32, 35:39)] %>%# Extract columns we wanna use
  filter(G > 8) %>%
  filter(QBR < QBR_q3 | Average.Salary > 2e+07) %>%
  filter(QBR > QBR_q1 | Average.Salary < 2e+07) %>%
  filter(Rate < Rtg_q3 | Average.Salary > 2e+07) %>%
  filter(Rate > Rtg_q1 | Average.Salary < 2e+07)


# Train a dummy model on all predictors just to get importance graph
set.seed(12345)
dummy_model <- randomForest(
  formula = Average.Salary ~ .,
  data = possible_predictors,
  ntree = 1000, # Arbitrarily high number of trees for importance certainty
  mtry = 4      # This number probably matters lmao
)

nrow(possible_predictors)
varImpPlot(dummy_model)
apply(importance(dummy_model), 2, function(x) sort(x, decreasing=TRUE))
plot(dummy_model)
```

Select most important predictors and train first model
- Note: I've only selected predictors that are easily obtainable
- Advanced Stats like Any.A (Adjusted Net Yards per Pass attempt) are good predictors but they're generally hard to obtain
- Depending on our deliverable scope, we could add them in later.
- I'm including QBR and Rating for advanced statistics because they're more common
```{r}
set.seed(12345)
span_model <- randomForest(
  formula = Average.Salary ~ QBR + Rate + Y.G_pass + Att_pass + TD_pass + Y.G_rush,
  data = possible_predictors,
  ntree = 200, # let's mess with this value later. The higher this gets, the more prone to overfitting
  mtry = 3 # Tested with 3 and 4, basically the same
)

span_model$mse[200]
span_model$rsq[200]

predictions <- as.vector(predict(span_model, newdata = QB_Data))
res <- data.frame(
  Name = QB_Data$Player,
  Year = QB_Data$Year,
  Salary = QB_Data$Average.Salary,
  Predicted_Salary = predictions,
  Diff = predictions - QB_Data$Average.Salary,
  PDiff = predictions / QB_Data$Average.Salary
)

# Some testers
res %>% arrange(desc(Predicted_Salary))
res %>% arrange(desc(Predicted_Salary)) %>% filter(Year == 2024)
res %>% arrange(desc(Year)) %>% filter(Name == "Brock Purdy")
```
Let's see how well it predicts contract year AAV
```{r}
Contract_year_datapoints <- QB_Data %>% filter(Year == Start.Year-1)
predictions <- as.vector(predict(span_model, newdata = Contract_year_datapoints))
res <- data.frame(
  Name = Contract_year_datapoints$Player,
  Year = Contract_year_datapoints$Year,
  Salary = Contract_year_datapoints$Average.Salary,
  Predicted_Salary = predictions,
  Diff = predictions - Contract_year_datapoints$Average.Salary,
  PDiff = predictions / Contract_year_datapoints$Average.Salary
)
res %>% arrange(desc(Predicted_Salary))
```

Test outputs with manually inserted data
```{r}
new_data <- data.frame(
  Player = "D'Brickshaw Ferguson",
  QBR = 74,
  Rate = 81,
  Y.G_pass = 5,
  Att_pass = 24,
  TD_pass = 8,
  Y.G_rush = 20,
  Average.Salary = 2
)

predictions <- as.vector(predict(span_model, new_data))
res <- data.frame(
  Name = new_data$Player,
  Salary = new_data$Average.Salary,
  Predicted_Salary = predictions,
  Diff = predictions - new_data$Average.Salary,
  PDiff = predictions / new_data$Average.Salary
)
res %>% arrange(desc(Predicted_Salary))
```

Potential improvements:
- Historical contract data via OTC
- Average over years to contract -> contract value
- Use XGBoost or a different regression model
  - I think XGBoost will be better, I have no idea truthfully
