---
title: "QB Salary Prediction"
output: html_document
date: "2025-03-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Packages and Data Import
```{r}
# Packages, install if necessary
library(dplyr)
library(randomForest)
library(ggplot2)
library(ggrepel)
library(gridExtra)

# Straight from github repo, assumes all files in same directory
Salary_Data <- read.csv("./NFL Salary Data.csv")
Receiving_Data <- read.csv("./Receiving Data.csv")
Rushing_Data <- read.csv("./Rushing Data.csv")
```

Formatting Data
```{r}
# Format Player_Data
# Remove space in team name and format dollar amounts into integers
Salary_Data <- Salary_Data %>% 
  mutate(Team.Currently.With = gsub("\\n.+", "", Team.Currently.With)) %>%
  mutate(
    Value = as.numeric(gsub("[$,]", "", Value)),
    Average.Salary = as.numeric(gsub("[$,]", "", Average.Salary)),
    Practical.Guarantee = as.numeric(gsub("[$,]", "", Practical.Guarantee)),
    Guarantee.at.Sign = as.numeric(gsub("[$,]", "", Guarantee.at.Sign))
  )

Receiver_Salary <- Salary_Data %>%
  filter(Pos == "WR" | Pos == "TE") %>% 
  select(Player, Average.Salary, Age.At.Signing, Start.Year) # Change Average.Salary if we want to use total contract or years

Receiver_Rushing_Data <- Rushing_Data %>% 
  filter(Pos == "WR" | Pos == "TE") %>%
  select(Player, Year, Att, Yds, TD, Y.G)

Receiver_Data <- Receiving_Data %>%
  left_join(Receiver_Rushing_Data, by= c("Player","Year"), suffix = c("_receive", "_rush")) %>% # Add in rushing data
  left_join(Receiver_Salary, by = "Player") %>% # Add in Salary Data
  filter(!is.na(Average.Salary)) %>%
  filter(R.G > 2)

# Replace NAs
Receiver_Data[is.na(Receiver_Data)] <- 0

# Remove duplicates
Receiver_Data <- Receiver_Data[!duplicated(Receiver_Data[c("Player", "Year")]), ]


nrow(Receiver_Data)
```

# Exploratory Data Analysis

Outlier Bounds via Quantiles
```{r warning=F} 
# Outlier testing
pd2024 <- Receiver_Data

quantilePlot <- function(df, col, title = col) {
  Q1 <- quantile(df[[col]], .25)
  Median <- quantile(df[[col]], .5)
  Q3 <- quantile(df[[col]], .75)
  IQR <- Q3 - Q1
  lower <- Q1 - 1.5 * IQR
  upper <- Q3 + 1.5 * IQR
  
  return(
    ggplot(df, aes(x = 0, y=.data[[col]])) +
    geom_boxplot() +
    geom_label(data = df, aes(y = Q1, label = paste("Q1:", round(Q1, 2)))) +
    geom_label(data = df, aes(y = Median, label = paste("Median:", round(Median, 2)))) +
    geom_label(data = df, aes(y = Q3, label = paste("Q3:", round(Q3, 2)))) +
    geom_text_repel(aes(label = 
      ifelse(
        .data[[col]] < lower | .data[[col]] > upper,
        paste(gsub(".+ ", "", Player), Year), ""
      )
    )) +
    labs(x = "", y = "", title = title)
  )
}


grid.arrange(
  quantilePlot(pd2024, "Y.G_receive", "Receiving Yards/Game"),
  quantilePlot(pd2024, "R.G", "Receptions/Game"),
  quantilePlot(pd2024, "Tgt", "Total Targets"),
  quantilePlot(pd2024, "TD_receive", "Touchdown Receptions"),
  nrow=2, ncol=2, top = "Performance metric Quantiles:"
)
```

Outlier Visualization based on Bounds
```{r}
salary_threshold <- mean(pd2024$Average.Salary)

# Contract v Age test
ggplot(Receiver_Salary, aes(Age.At.Signing, Average.Salary)) + geom_point() +
  # geom_text_repel(aes(label = Player)) +  # idt we neeeed names here
  theme(legend.position = "none") +
  labs(
    title = "How Age affects Contract Value",
    y="Yearly Salary signed for",
    x = "Age at Contract Signing"
  )

# QBR Outlier Test
ggplot(pd2024, aes(Y.G_receive, Average.Salary, color=Player)) + geom_point() +
  geom_text_repel(aes(label = 
    ifelse(
      (Y.G_receive > quantile(Y.G_receive, .75) & Average.Salary < salary_threshold) | (Y.G_receive < quantile(Y.G_receive, .25) & Average.Salary > salary_threshold),
      paste(gsub("[^A-Z]+", ".", Player), sub("20", "", paste(Year))),
      ""
    ) 
  )) +
  labs(
    title = "Yards/Game vs Average Salary, Overperformers and Underperformers labeled",
    y = "Average Salary"
  ) + 
  theme(legend.position = "none")

# Passer Rating outlier test
ggplot(pd2024, aes(R.G, Average.Salary, color=Player)) + geom_point() +
  geom_text_repel(aes(label =
    ifelse(
      (R.G > quantile(R.G, .75) & Average.Salary < salary_threshold) | (R.G < quantile(R.G, .25) & Average.Salary > salary_threshold),
      paste(gsub("[^A-Z]+", ".", Player), sub("20", "", paste(Year))),
      ""
    )
  )) +
  labs(
    title = "Receptions/Game vs Average Salary, Overperformers and Underperformers labeled",
    y = "Average Salary",
    x = "Receptions/Game"
  ) + 
  theme(legend.position = "none")

# Y.G Outlier test
ggplot(pd2024, aes(Tgt, Average.Salary, color=Player)) + geom_point() +
  geom_text_repel(aes(label = 
    ifelse(
      (Tgt > quantile(Tgt, .75) & Average.Salary < salary_threshold) | (Tgt < quantile(Tgt, .25) & Average.Salary > salary_threshold),  
      paste(gsub("[^A-Z]+", ".", Player), sub("20", "", paste(Year))),
      ""
    ) 
  )) +
  labs(
    title = "Targets v Average Salary, Overperformers and Underperformers labeled",
    y = "Average Salary",
    x = "Total Targets"
  ) + 
  theme(legend.position = "none")

# Total Touchdown test
ggplot(pd2024, aes(TD_receive, Average.Salary, color=Player)) + geom_point() +
  geom_text_repel(aes(label = 
    ifelse(
      (TD_receive > quantile(TD_receive, .75) & Average.Salary < salary_threshold) | (TD_receive < quantile(TD_receive, .25) & Average.Salary > salary_threshold),  
      paste(gsub("[^A-Z]+", ".", Player), sub("20", "", paste(Year))),
      ""
    ) 
  )) +
  labs(
    title = "Receiving TD v Average Salary, Overperformers and Underperformers labeled",
    y = "Average Salary",
    x = "Receiving Touchdowns"
  ) + 
  theme(legend.position = "none")
```


# Training the Model

Training Data Generation
```{r}
# Use these to get static values to remove outliers with
YG_q1 <- quantile(Receiver_Data$Y.G_receive , .25)
YG_q3 <- quantile(Receiver_Data$Y.G_receive , .75)
RG_q1 <- quantile(Receiver_Data$R.G, .25)
RG_q3 <- quantile(Receiver_Data$R.G, .75)


possible_predictors <- Receiver_Data[,c(3, 5:20, 22:29)] %>% # Extract columns we wanna use
 filter(Y.G_receive < YG_q3 | Average.Salary > mean(Average.Salary)) %>%
 filter(Y.G_receive > YG_q1 | Average.Salary < mean(Average.Salary)) %>%
 filter(R.G < RG_q3 | Average.Salary > mean(Average.Salary)) %>%
 filter(R.G > RG_q1 | Average.Salary < mean(Average.Salary))


# Train a dummy model on all predictors just to get importance graph
set.seed(12345)
dummy_model <- randomForest(
  formula = Average.Salary ~ .,
  data = possible_predictors,
  ntree = 1000, # Arbitrarily high number of trees for importance certainty
  mtry = 4      # This number probably matters lmao
)

nrow(possible_predictors)
varImpPlot(dummy_model)
apply(importance(dummy_model), 2, function(x) sort(x, decreasing=TRUE))
plot(dummy_model)
```

Select most important predictors and train first model
```{r}
set.seed(12345)
span_model <- randomForest(
  formula = Average.Salary ~ Y.G_receive + R.G + Tgt + TD_receive,
  data = possible_predictors,
  ntree = 200, # let's mess with this value later. The higher this gets, the more prone to overfitting
  mtry = 3 # Tested with 3 and 4, basically the same
)

span_model$mse[200]
span_model$rsq[200]

predictions <- as.vector(predict(span_model, newdata = Receiver_Data))
res <- data.frame(
  Name = Receiver_Data$Player,
  Year = Receiver_Data$Year,
  Salary = Receiver_Data$Average.Salary,
  Predicted_Salary = predictions,
  Diff = predictions - Receiver_Data$Average.Salary,
  PDiff = predictions / Receiver_Data$Average.Salary
)

# Some testers
res %>% arrange(desc(Predicted_Salary))
res %>% arrange(desc(Predicted_Salary)) %>% filter(Year == 2024)
```

Visualizing the difference
```{r}
ggplot(res %>% filter(Year == 2023) %>% arrange(desc(Predicted_Salary)) %>% slice_head(n=30), 
       aes(x = Salary, xend = Predicted_Salary, y = reorder(Name, Predicted_Salary))) +
  geom_segment(aes(yend = Name)) +
  geom_point(aes(y = Name, x = Salary), color="red") + 
  geom_point(aes(y = Name, x = Predicted_Salary), color="blue")
```

Let's see how well it predicts contract year AAV
```{r}
Contract_year_datapoints <- Receiver_Data %>% filter(Year == Start.Year-1)
predictions <- as.vector(predict(span_model, newdata = Contract_year_datapoints))
res <- data.frame(
  Name = Contract_year_datapoints$Player,
  Year = Contract_year_datapoints$Year,
  Salary = Contract_year_datapoints$Average.Salary,
  Predicted_Salary = predictions,
  Diff = predictions - Contract_year_datapoints$Average.Salary,
  PDiff = predictions / Contract_year_datapoints$Average.Salary
)
res %>% arrange(desc(Predicted_Salary))
```

Test outputs with manually inserted data
```{r}
new_data <- data.frame(
  Player = "D'Brickshaw Ferguson",
  QBR = 74,
  Rate = 100,
  Y.G_pass = 265,
  Att_pass = 450,
  TD_pass = 25,
  Y.G_rush = 20,
  Average.Salary = 2
)

predictions <- as.vector(predict(span_model, new_data))
res <- data.frame(
  Name = new_data$Player,
  Salary = new_data$Average.Salary,
  Predicted_Salary = predictions,
  Diff = predictions - new_data$Average.Salary,
  PDiff = predictions / new_data$Average.Salary
)
res %>% arrange(desc(Predicted_Salary))
```

Potential improvements:
- Historical contract data via OTC
- Average over years to contract -> contract value
- Use XGBoost or a different regression model
  - I think XGBoost will be better, I have no idea truthfully
